probKanren: A Simple Probabilistic extension for microKanren

### 2021-09-15 paper/main.tex
Fix SMC bits

This probKanren program draws 1000 samples from a conditional normal distribution.

@@

To create \textsc{probKanren} we extend this grammar with distribution clauses
such as \texttt{normal} and \texttt{bern}. These are just another type
of goal expression. We do not need to add a notion of probabilistic
variables as they can just be seen as logic variables constrained in
a particular way.

@@

Sequential Monte Carlo\cite{chopin2020introduction}(SMC) is an efficient
online way to sample from probabilistic models especially suited for
state-space domains. If we imagine our probabilistic programs as
straight-line programs with no control-flow we can imagine numbering
every sample function $f_1, f_2, \ldots, f_n$ and every observe
function $g_1, g_2, \ldots, g_n$ then our probability density over our
random variables $x$ and observed data $y$ can be defined as:

\begin{equation}
  p(x, y) = \prod_i f(x_i \mid x_{0:n-1})g(y_i \mid x_{0:n})
\end{equation}

@@

If we imagine running this program each execution trace can be seen as
a sample from the distribution. If we then weigh by their likelihood
as we run them that collection of execution traces will be an empirical
distribution of the given program. We call each of these execution traces
particles and the below method is how we obtain them along with their weights $w$.

\begin{eqnarray}
  x_n &\sim& f(x_n \mid x_{0:n-1}) \\
  w_n^{(k)} &=& \frac{g(y_n \mid x_{0:n})f(x_n \mid x_{0:n-1})}{q(x_n \mid x_{0:n-1})} \\
  W_n^{(k)} &=& W_{n-1}^{(k)} w_n^{(k)}
\end{eqnarray}

The weights are then normalised at the end to make a proper probability distribution.

@@

@@

As the SMC resampling step was defined in the previous section
Particle Cascades \cite{PaigeWDT14} remove this barrier allowing
every particle to be resampled asynchronously with the associated
weights being relative to a global running average. % double-check this

@@

An initial set of particles is created from the probabilistic program
when it is first run. As disjunctions (\texttt{disj}) are encountered,
we split evenly the number of particles allocated to each stream. Whenever
we encounter a unification primitive, we run a resampling step. This
helps to prune low-weight particles and replicate high weight ones.

@@

@@

\section{Semantics of probKanren}

>>>>

In the above program, we have a probabilistic model for the Gaussian
unknown mean problem. We observe a value 4 from the $\mathcal{N}(q,2)$
distribution, and that $q$ has a prior distribution of $\mathcal{N}(0, 3)$.
This \textsc{probKanren} program draws 1000 samples from a conditional normal distribution
that represents the posterior probability distribution associated with $q$.

@@

To create \textsc{probKanren} we extend this grammar with distribution
clauses such as \texttt{normal} and \texttt{bern}. Distribution clauses
take as arguments the parameters of the distribution and a last argument
representing a draw from that distribution. For example, in \texttt{(normal 0 1 x)}
means \texttt{x} represents a draw from $\mathcal{N}(0,1)$ distribution.

These are just another type of goal expression. We do not need to add
a notion of probabilistic variables to the language as they can
treated as logic variables constrained in a particular way. The
semantics of the language though does change from a set of answers to
a query to a probability distribution of answers to a query.

@@

Sequential Monte Carlo\cite{chopin2020introduction}(SMC) is an efficient
online way to sample from probabilistic models especially suited for
state-space domains. If we imagine our probabilistic programs as
straight-line programs with no control-flow we can imagine numbering
every sample function $f_1, f_2, \ldots, f_n$ and every observe
function $g_1, g_2, \ldots, g_n$ then our probability density over our
random variables $x_{0:T}$ and observed data $y_{0:T}$ can be defined as:

\begin{equation}
  p(x_{0:T}, y_{0:T}) = p(x_0)\prod_{t=1}^T f_t(x_t \mid x_{t-1})\prod_{t=0}^T g_t(y_t \mid x_t)
\end{equation}

@@

The simplest way to sample from our target distribution $p(x_{0:T}
\mid y_{0:T})$ is to samples from a sequence of intermediary
distributions $p(x_{0:t} \mid y_{0:t})$ where $t$ goes from $1$ to
$T$. We do this by sampling a population of $N$ particles
$\{x^{(1)}_t, \ldots, x^{(i)}_t, \ldots x^{(N)}_t\}$ from a proposal
distribution $q(x_t \mid x_{0:t-1}, y_{0:t})$, which when combined
with a set of importance weights $\{w^{(1)}_t, \ldots, w^{(i)}_t,
\ldots w^{(N)}_t\}$ lets us approximate each intermediary
distribution.

Thanks to the recurrence relation

\begin{equation}
p(x_{0:t} \mid y_{0:t}) = p(x_{0:t-1} \mid y_{0:t-1}) \frac{f_t(x_t \mid x_{t-1}) g_t(y_t \mid x_t)}{p(y_t \mid y_{0:t-1})}
\end{equation}

We know in $T$ rounds we compute the desired distribution as follows:

Initialise with:

\begin{align*}
x_0^{(i)} &\sim p(x_0) \\
w_0^{(i)} &= 1/N
\end{align*}

Then for $t$ from $1$ to $T$

\begin{align*}
  x_t^{(i)} &\sim q(x_t \mid x^{(i)}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{g(y_t \mid x^{(i)}_t)f(x^{(i)}_t \mid x^{(i)}_{t-1})}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align*}

We get the best results if $q(x_t \mid x_{0:t-1})$ is equal to $p(x_t \mid x_{0:t-1}, y_{0:t})$.

@@

Unfortunately, SIS is likely to over time lead to most of the weight
in a small fraction of the particles with the rest of the particles
having negligible weight. This is sometimes called \emph{weight degeneracy}. We
address this by replicating the particles with high weight and
dropping the ones with low weight. This is called a resampling step,
and it happens after each round.

\begin{align*}
  a_t^{(i)} &\sim r(w_{t-1}^{(1:N)}) \\
  x_t^{(i)} &\sim q(x_t \mid x^{(a_t^{(i)})}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{g(y_t \mid x^{(i)}_t)f(x^{(i)}_t \mid x^{(i)}_{t-1})}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align*}

@@

Unfortunately, SMC as defined here requires having access to all the
particles at every step in the process. This conflicts with the
functional nature of our implementation. Instead we make use of
Particle Cascades \cite{PaigeWDT14}, remove this barrier allowing every
particle to be resampled asynchronously with the associated weights
being relative to a global running average. % double-check this

@@

@@

An initial set of particles is created from the probabilistic program
when it is first run. When primitives such as \texttt{normal} we first
check is all the parameter terms are grounded and then if the last
argument is fresh, we sample a value for it for each particle and then
add this sampled value along with the associated logic variable to the
substitution associated with that particle. If all the terms are ground,
we treat the primitive like an observation statement and multiply the
weights of each substitution with the likelihood of the observation.

As disjunctions (\texttt{disj}) are encountered, we split evenly the
number of particles allocated to each stream. Whenever we encounter a
unification primitive, we run a resampling step. This helps to prune
low-weight particles and replicate high weight ones.

@@
